Automated neuroimaging data management with the Sea filesystem

### 2022-01-25 paper/sea-neuro/paper_neuro.tex
paper updates

\algnewcommand\algorithmicforeach{\textbf{for each}}
\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}
\algblock{Input}{EndInput}
\algnotext{EndInput}
\algblock{Output}{EndOutput}
\algnotext{EndOutput}
\newcommand{\tristan}[1]{\color{orange}\textbf{From Tristan:} #1\color{black}\xspace}
\newcommand{\Desc}[2]{\State \makebox[2em][l]{#1}#2}
\newcommand{\todo}[1]{\marginpar{\parbox{18mm}{\flushleft\tiny\color{red}\textbf{TODO}: #1}}}

@@

    Neuroimaging workflows, particularly preprocessing ones, may lead to a magnification in output sizes due to the presence of multiple intermediary processing steps. They also produce
    significant amounts of intermediary data due to being composed of a multitude of steps. 
\end{abstract}
}

@@

HPC systems rely on scalable network-based parallel filesystems (e.g. Lustre) for storage. While such file
systems offer excellent performance, they are shared between all the users on the cluster. Meaning that users
with data-intensive analyses can effectively deteriorate the performance of the shared file system for all users on
the cluster. Solutions for improving shared file system performance include throttling the data intensive workloads
or recommending the use of Burst Buffers~\cite{bb} (e.g. reserving a compute node for storage or leveraging
local compute storage during processing). The latter, however, requiring that the user manages their data to
and from the Burst Buffer.

@@

As neuroimaging pipeline results typically require post-processing and HPC compute-local resources are only accessible during the
reserved duration, Sea provides functionality to flush and evict data to persistent shared storage. This is accomplished via a separate
thread (known as the ``flusher'') that moves data from the caches to long-term storage. A separate thread is used, in this case, to avoid
interrupting on going processing with data management operations.
Users must inform Sea of files that need to be
persisted to storage within a file called \texttt{.sea\_flushlist}, and temporary files which can be removed from cache within a file
called the \texttt{.sea\_evictlist}. Both these files can be populated using regular expressions to denote the paths to flush and evict.
If a file occurs in both the \texttt{.sea\_flushlist} and \texttt{.sea\_evictlist}, Sea will interpret this as a \texttt{move} operation
and simply move the file from local to persistent storage rather than copying it. Files that will be reused by the application should only
be flushed rather than flushed and evicted, as files can benefit from speedups related to reading from fast rather than slow storage.
Sea currently also provides a rudimentary prefetch thread that can move files located within the Sea filesystem to the fastest available cache.
To use Sea's prefetch capabilities, a file called \texttt{.sea\_prefetchlist} needs to be populated using regular expressions like the
flushing and eviction files.

@@

\subsubsection{Datasets}
Functional magnetic resonance imaging datasets can vary greatly in total number
of images and number of volumes within each image. To adequately capture the 
diversity of datasets and when Sea could be pertinent, we selected three datasets of varying
sizes: 1) OpenNeuro's ds001545 dataset~\cite{ds001545}, 2) the PREVENT-AD  dataset~\cite{preventad}, and
3) the HCP dataset~\cite{hcp}. The ds001545 dataset is a total of 45.94~$GB$ (1778 files) and consists of data collected
from 30 subjects in a single session. The PREVENT-AD dataset is a 255.0~$GB$ dataset (53061 files) consisting of data collected
from 308 subjects. The HCP dataset is the largest of the three datasets at 85.4~$TB$ (x files) consisting of data collected from
1113 subjects.

@@

Similarly to datasets, preprocessing pipelines can also vary in duration as a result of methodological differences.
Thus, we preprocessed each dataset using four standard preprocessing pipelines: FSL~\cite{fsl}, AFNI~\cite{AFNI},
SPM~\cite{SPM}, fMRIPrep~\cite{fmriprep}. Table~\ref{tb1} provides an reference of the differences in computation and data-intensivity
of the different pipelines on a single subject of the ds001545 dataset.

Each tool was set to only run the preprocessing pipeline. Scripts used to execute each pipeline can be found at~\ref{github.com}.
For the FSL preprocessing pipeline, the default settings were maintained with the exception of slice-timing correction (interleaved),
intensity normalization and non-linear registration to standard space.

%% Add SPM here

@@

\subsection{Controls}
\subsection{Infrastructure}

\section{Results}

\section{Discussion}
talk about testing

>>>>

\algnewcommand\algorithmicforeach{\textbf{for each}}
\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}
\algblock{Input}{EndInput}
\algnotext{EndInput}
\algblock{Output}{EndOutput}
\algnotext{EndOutput}
\newcommand{\Desc}[2]{\State \makebox[2em][l]{#1}#2}
\newcommand{\todo}[1]{\marginpar{\parbox{18mm}{\flushleft\tiny\color{red}\textbf{TODO}: #1}}}

@@

    Neuroimaging workflows, particularly preprocessing ones, may lead to a magnification in output sizes due to <add reason>. They also produce
    significant amounts of intermediary data due to being composed of a multitude of steps. 
\end{abstract}
}

@@

HPC systems rely on scalable network-based parallel filesystems (e.g. Lustre) for storage. While such file
systems offer excellent performance, they are shared between all the users on the cluster. Meaning that users
with data-intensive can effectively deteriorate the performance of the shared file system for all users on
the cluster. Solutions for improving shared file system performance include throttling the data intensive workloads
or recommending the use of Burst Buffers~\cite{bb} (e.g. reserving a compute node for storage or leveraging
local compute storage during processing). The latter, however, requiring that the user manages their data to
and from the Burst Buffer.

@@

As neuroimaging pipeline results are typically required post-processing and HPC compute-local resources are only accessible during the
reserved duration, Sea provides functionality to flush and evict data to persistent shared storage. This is accomplished via a separate
thread (known as the ``flusher'') that moves data from the caches to long-term storage. A separate thread is used, in this case, to avoid
interrupting on going processing with data management operations.
Users must inform Sea of files that need to be
persisted to storage within a file called \texttt{.sea\_flushlist}, and temporary files which can be removed from cache within a file
called the \texttt{.sea\_evictlist}. Both these files can be populated using regular expressions to denote the paths to flush and evict.
If a file occurs in both the \texttt{.sea\_flushlist} and \texttt{.sea\_evictlist}, Sea will interpret this as a \texttt{move} operation
and simply move the file from local to persistent storage rather than copying it. Files that will be reused by the application should only
be flushed rather than flushed and evicted, as files can benefit from speedups related to reading from fast rather than slow storage.
Sea currently also provides a rudimentary prefetch thread that can move files located within the Sea filesystem to the fastest available cache.
To use Sea's prefetch capabilities, a file called \texttt{.sea\_prefetchlist} needs to be populated using regular expressions like the
flushing and eviction files.

@@

\subsubsection{Datasets}
fMRI datasets may vary greatly in total number
of images and number of volumes within each image. To adequately capture the 
diversity of datasets and the applicability of Sea, we have selected three datasets of varying
time and space resolutions (Table~\ref{table:data}): 1) OpenNeuro's ds001545 dataset~\cite{ds001545},
2) the PREVENT-AD dataset~\cite{preventad}, and 3) the HCP dataset~\cite{hcp}.
The ds001545 dataset is a total of 45.94~$GB$ (1778 files) and consists of data collected
from 30 participants in a single session watching 3 different clips (i.e. three runs/participants) of the The Grand Budapest Hotel.

The PREVENT-AD dataset is publicly available dataset comprising of data from 330 participants with a familial history
of Alzheimer's Disease.
At the time of the experiments, the dataset contained 255.0~$GB$ (53061 files) originating from the 308 subjects available on DataLad~\cite{}.

The HCP project, whose aim is to characterize brain connectivity using data collected from 1200 subjects, is the largest of the three datasets.
At the time of our experimments, the dataset obtained consisted 85.4~$TB$ (x files) of data collected from 1113 subjects.

By experimenting with datasets of different spatial and temporal resolutions, the amount of compute may be affected.
Naturally, with more data there may inevitably be more computations performed, but also, the code itself may
chose to perform entirely different computations as a result of differences in resolution. Moreover, parallelism and program I/O behaviour
may be impacted depending on the particular characteristics of the data. Capturing this potential
variety in processing will provide us with a better understanding of when and where Sea can be useful.

\begin{table*}[t]
  \small\centering
\begin{tabular}{|c c c c c|}
  \hline
  Dataset & Total Size (MB) & Total Number of images & Number of images processed & Total size of processed (MB) \\
  \hline
  \multirow{3}{*}{PREVENT-AD} & \multirow{3}{*}{289532} & \multirow{3}{*}{53061} & 1 & 51.26\\
  & & & 8 & \\
  & & & 16 & \\
  \hline
  \multirow{3}{*}{ds001545} & \multirow{3}{*}{27377} & \multirow{3}{*}{1778} & 1 & 280.27\\
  & & & 8 & \\
  & & & 16 & \\
  \hline
  \multirow{3}{*}{HCP} & \multirow{3}{*}{83140079} & \multirow{3}{*}{X} & 1 &  1299.93\\
  & & & 8 & \\
  & & & 16 & \\

  \hline

  \hline
\end{tabular}\caption{Dataset characteristics}\label{table:data}
\end{table*}

@@

Similarly to datasets, preprocessing pipelines can also vary in duration as a result of methodological differences.
Thus, we preprocessed each dataset using four standard preprocessing pipelines: FSL~\cite{fsl}, AFNI~\cite{AFNI},
SPM~\cite{SPM}, fMRIPrep~\cite{fmriprep}. Table~\ref{table:pipelines} provides an reference of the differences in computation and data-intensivity
of the different pipelines on a single subject of all three of the datasets.

Each tool was set to only run the preprocessing pipeline. Scripts used to execute each pipeline can be found at~\ref{github.com/valhayot/sea-slurm-scripts}.

For the FSL preprocessing pipeline, the default settings were maintained with the exception of slice-timing correction (interleaved),
intensity normalization and non-linear registration to standard space.

%% Add SPM here
To preprocess the datasets with SPM12, we reused the template described in \cite{haitas2021}, with fieldmap correction removed.
As the template was designed for a specific dataset, we updated the paths to the fMRI and anatomical images,
as well as the dimensions, such as the number of volumes and slices and the TR and TA, for each dataset. Furthermore,
we set the slice order to interleaved in all cases.

@@

\begin{table*}[t]
  \small\centering
\begin{tabular}{|c c c c c c|}
  \hline
  Tool & Dataset &  Output Size (MB) & Total libc calls & Libc Lustre calls & Compute time (s) \\
  \hline
  \multirow{3}{1em}{AFNI} & PREVENT-AD & 539.6 & 272390 & 4348 & 103.25 \\
  & ds001545 & 3063.4 & 281707 & 4570  & 280.30 \\
  & HCP & 18720.4 & 305310 & 5367 & 816.16 \\
  \hline
  \multirow{3}{1em}{FSL Feat} & PREVENT-AD & 253.7 & &  & 1338.29 \\
  & ds001545 & 550.6 & &  & 2145.96 \\
  & HCP & 1607.9 & & & 6596.46 \\
  \hline
  \multirow{3}{1em}{SPM} & PREVENT-AD & 331.0 & 34868 & 10048 &  483.67 \\
  & ds001545 & 744.4 & 47251 & 19568 & 446.53 \\
  & HCP & 2082.5 & 54753 & 25226 & 715.43 \\

  \hline
\end{tabular}\caption{Pipeline execution characteristics}\label{table:pipelines}
\end{table*}
\subsection{Controls}
Our objective is to measure the speedup brought by Sea on real neuroimaging pipelines.
Although Sea is agnostic to the internals of the pipelines and their output results, incorrect parameter
selection and resulting outputs may result in I/O patterns that would not necessarily be observed had the
pipeline been executed properly.

Visual QC on the pipeline to ensure it was correctly registered and skull-striped to the template brain (MNI template).

To validate that the outputs produced were sufficiently correct, we overlayed the output on the input image.
This allowed up to see if steps such as skull-stripping and registration were correctly performed on the data.
We repeated this step for all datasets and all pipelines processing a single image. In addition, we compared output
number of files and output size to ensure that the pipeline was not producing different outputs depending on the file
system used.

% compare md5sums of uncompressed files.

\subsection{Infrastructure}

Our experiments were exected on two different clusters: 1) a local cluster and 2) Compute Canada's Beluga cluster.
The local cluster comprised of 8 compute nodes with 256GB memory and 125GB tmpfs, running Centos 8. Storage on the local
cluster was comprised of 4 Lustre ZFS storage nodes with 44 HDD OSTs and 1 MDS with a single MDT. The
storage nodes were connected to the compute nodes via a 20Gbps ethernet connection.
The cluster's hardware is similar to that of Beluga.

The Beluga cluster

%Busy writers
\section{Results}

\section{Discussion}
\subsection{Significance of speedups with Sea}

\subsection{Other use cases}
writes a lot but it's actually more complicated. When dataset increases
\subsection{Predicting speedups}
 difficulties with summarizing speedups into guidelines.
 variability is important and it's coming from the interaction between infrastructure and application.
 Benefits of sea depend on parameters such as load of the filesystem which is multifold and the application compute and
 file system configuration. 
 page cache
 examples:
  - large dataset - > more computation
  - data access patterns and bursty-ness
  - small datasets + metadata accesses

\subsection{Libc interception of neuroimaging pipelines}

As demonstrated by our Sea experiments, we have found that libc interception is effective at capturing all file system
calls executed by the applications. This means that Sea is compatible with the most prominently used neuroimaging
preprocessing pipelines. Although we have not extensively against all possible parameters and toolbox applications, we
assume that use of glibc within the toolboxes themselves are likely to be consistent.

The ability to use libc interception over more common alternative methods such as kernel-based or fuse-based filesystems
was essential in ensuring ease-of-use and minimal performance overheads.

opportunities for other optimizations

Sea may be beneficial
\subsection{Should Sea always be used?}

user

sysadmin

write intensive

\subsection{Effective use of available cluster resources}
\subsection{Prefetching, flushing and eviction with Sea}
\subsection{Comparisons to best case scenario}

talk about testing
